\textbf{Efficient Implementation Strategies for Block Ciphers on ARMv8}\\ {\footnotesize Bachelorarbeit}

### 2023-05-30 Paper/chapters/implementation-strategies.tex
finalize

\section{Strategies for \texttt{GIFT}}

@@

Table-driven programming is a simple way to increase performance of operations
by tabulating the results, therefore requiring only a single memory access to
acquire the result. This approach is obviously limited to manageable table
sizes, so while tabulating a function like the AES S-box
$S_{AES}:\F{8}\rightarrow \F{8}$ requires only $2^{11}$ space,
tabulating the \texttt{GIFT} permutation layer
$P_{GIFT}:\F{64}\rightarrow \F{64}$ would require
$2^{70}$ space, which is totally unfeasible.

A common approach is to tabulate the output of each S-box, including the
diffusion layer, and then XORing the results together. Let $n$ denote the
internal cipher state size and $s$ the size of a single S-box in bits. For each
S-box $S_i,i\in\{0,\dots,\frac{n}{s}\}$, we can construct a mapping
$T_i:\F{s}\rightarrow \F{n}$ representing substitution with subsequent
permutation of that single S-box. The cipher state before round key addition is
then given by $\bigoplus_{i=0}^{\frac{n}{s}-1}{T_i(m_i)}$ for each $s$-bit
message chunk $m_i$. This approach requires space of
$\frac{n}{s}|\F{s}|n=\frac{n^2 2^s}{s}$ bits, which, for \texttt{GIFT-64},
results in a manageable size of $\frac{64^2 2^4}{4}=2^{14}$ bits which equals
$16$ KiB.

@@

For \texttt{GIFT-64}, table construction is relatively straightforward and can
be done as follows:

@@

The tables for \texttt{GIFT-128} can be generated in a similar way by looping
through all 32 S-boxes instead of 16 on line \ref{lst:tablesbox}.

@@

Bitsliced implementation techniques were first introduced to improve the
performance of DES in 1997 and work by viewing a processor with $n$-bit
registers as a machine capable of executing $n$ bitwise operations at
once \cite{desslicing:1997}. Bitslicing offers a performance advantage by
splitting up $n$ bits into $m$ slices to achieve a more efficient
representation which can exploit this bitwise parallelism. The structure of
\texttt{GIFT} naturally offers possibilities for bitslicing. We split the
cipher state bits $b_{63}b_{62}\dots b_0$ into four slices $S_i,
i\in\{0,1,2,3\}$ such that the $i$-th slice contains all $i$-th bits of the
individual S-boxes. This is equivalent to transposing the bit matrix.

@@

This representation offers multiple advantages. We first note that computation
of the S-box can be executed in parallel, similar to the \texttt{vperm}
technique above. This can be done by finding a bitwise instruction sequence to
apply the S-box which has already been proposed by the original \texttt{GIFT}
authors:

@@

In the case of \texttt{GIFT-64} with $BS=64$ and ARM NEON with $RS=128$, we get

@@

The original paper proposes various platform-independent ways to implement
Camellia efficiently. Only some of these apply to the ARMv8 architecture since
features like the inline barrel shifter and bitfield manipulation instructions
generally offer better performance.

@@

        \begin{alignat*}{9}
            & SP_0(y_{0(8)})=(&s_0(y_0),&& s_0(y_0),&& s_0(y_0),&& 0,&& s_0(y_0),&& 0,&& 0,&& s_0(y_0)&) \\
            & SP_1(y_{1(8)})=(&0,&& s_1(y_1),&& s_1(y_1),&& s_1(y_1),&& s_1(y_1),&& s_1(y_1),&& 0,&& 0&) \\
            & SP_2(y_{2(8)})=(&s_2(y_2),&& 0,&& s_2(y_2),&& s_2(y_2),&& 0,&& s_2(y_2),&& s_2(y_2),&& 0&) \\
            & SP_3(y_{3(8)})=(&s_3(y_3),&& s_3(y_3),&& 0,&& s_3(y_3),&& 0,&& 0,&& s_3(y_3),&& s_3(y_3)&) \\
            & SP_4(y_{4(8)})=(&0,&& s_2(y_4),&& s_2(y_4),&& s_2(y_4),&& 0,&& s_2(y_4),&& s_2(y_4),&& s_2(y_4)&) \\
            & SP_5(y_{5(8)})=(&s_3(y_5),&& 0,&& s_3(y_5),&& s_3(y_5),&& s_3(y_5),&& 0,&& s_3(y_5),&& s_3(y_5)&) \\
            & SP_6(y_{6(8)})=(&s_4(y_6),&& s_4(y_6),&& 0,&& s_4(y_6),&& s_4(y_6),&& s_4(y_6),&& 0,&& s_4(y_6)&) \\
            & SP_7(y_{7(8)})=(&s_1(y_7),&& s_1(y_7),&& s_1(y_7),&& 0,&& s_1(y_7),&& s_1(y_7),&& s_1(y_7),&& 0&)
        \end{alignat*}

        , we can compute $(P\circ S)(X_{(64)})$ using only 8 table lookups and 7
        XORs in the following way:

>>>>

\section{Strategies for GIFT}

@@

Table-driven programming is a simple way to increase performance of operations
by tabulating the results, therefore requiring only a single memory access to
acquire the result. This approach is obviously limited to manageable table
sizes, so while tabulating a function like the AES S-box
$S_{AES}:\F{8}\rightarrow \F{8}$ requires only $2^{11}$ space, tabulating the
GIFT permutation layer $P_{GIFT}:\F{64}\rightarrow \F{64}$ would require
$2^{70}$ space, which is totally unfeasible.

A common approach is to tabulate the output of each S-box, including the
diffusion layer, and then XORing the results together. Let $n$ denote the
internal cipher state size and $s$ the size of a single S-box in bits. For each
S-box $S_i,i\in\{0,\dots,\frac{n}{s}\}$, we can construct a mapping
$T_i:\F{s}\rightarrow \F{n}$ representing substitution with subsequent
permutation of that single S-box. The cipher state before round key addition is
then given by $\bigoplus_{i=0}^{\frac{n}{s}-1}{T_i(m_i)}$ for each $s$-bit
message chunk $m_i$. This approach requires space of
$\frac{n}{s}|\F{s}|n=\frac{n^2 2^s}{s}$ bits, which, for GIFT-64,
results in a manageable size of $\frac{64^2 2^4}{4}=2^{14}$ bits which equals
$16$ KiB.

@@

For GIFT-64, table construction is relatively straightforward and can
be done as follows:

@@

The tables for GIFT-128 can be generated in a similar way by looping
through all 32 S-boxes instead of 16 on line \ref{lst:tablesbox}.

@@

Bitsliced implementation techniques were first introduced to improve the
performance of DES in 1997 and work by viewing a processor with $n$-bit
registers as a machine capable of executing $n$ bitwise operations at
once \cite{desslicing:1997}. Bitslicing offers a performance advantage by
splitting up $n$ bits into $m$ slices to achieve a more efficient
representation which can exploit this bitwise parallelism. The structure of
GIFT naturally offers possibilities for bitslicing. We split the
cipher state bits $b_{63}b_{62}\dots b_0$ into four slices $S_i,
i\in\{0,1,2,3\}$ such that the $i$-th slice contains all $i$-th bits of the
individual S-boxes. This is equivalent to transposing the bit matrix.

@@

This representation offers multiple advantages. We first note that computation
of the S-box can be executed in parallel, similar to the \texttt{vperm}
technique above. This can be done by finding a bitwise instruction sequence to
apply the S-box which has already been proposed by the original GIFT authors:

@@

In the case of GIFT-64 with $BS=64$ and ARM NEON with $RS=128$, we get

@@

The original paper proposes various platform-independent ways to implement
Camellia efficiently \cite{camellia:2001}. Only some of these apply to the
ARMv8 architecture since features like the inline barrel shifter and bitfield
manipulation instructions generally offer better performance.

@@

        \begin{alignat*}{9}
            & SP_0(y_{0(8)})=(&s_0(y_0),&& s_0(y_0),&& s_0(y_0),&& 0,&& s_0(y_0),&& 0,&& 0,&& s_0(y_0)&) \\
            & SP_1(y_{1(8)})=(&0,&& s_1(y_1),&& s_1(y_1),&& s_1(y_1),&& s_1(y_1),&& s_1(y_1),&& 0,&& 0&) \\
            & SP_2(y_{2(8)})=(&s_2(y_2),&& 0,&& s_2(y_2),&& s_2(y_2),&& 0,&& s_2(y_2),&& s_2(y_2),&& 0&) \\
            & SP_3(y_{3(8)})=(&s_3(y_3),&& s_3(y_3),&& 0,&& s_3(y_3),&& 0,&& 0,&& s_3(y_3),&& s_3(y_3)&) \\
            & SP_4(y_{4(8)})=(&0,&& s_2(y_4),&& s_2(y_4),&& s_2(y_4),&& 0,&& s_2(y_4),&& s_2(y_4),&& s_2(y_4)&) \\
            & SP_5(y_{5(8)})=(&s_3(y_5),&& 0,&& s_3(y_5),&& s_3(y_5),&& s_3(y_5),&& 0,&& s_3(y_5),&& s_3(y_5)&) \\
            & SP_6(y_{6(8)})=(&s_4(y_6),&& s_4(y_6),&& 0,&& s_4(y_6),&& s_4(y_6),&& s_4(y_6),&& 0,&& s_4(y_6)&) \\
            & SP_7(y_{7(8)})=(&s_1(y_7),&& s_1(y_7),&& s_1(y_7),&& 0,&& s_1(y_7),&& s_1(y_7),&& s_1(y_7),&& 0&)
        \end{alignat*},

        we can compute $(P\circ S)(X_{(64)})$ using only 8 table lookups and 7
        XORs in the following way:

### 2023-05-30 Paper/chapters/implementation.tex
finish draft
finalize

Implementations in the C programming language for the presented strategies can
be found in Appendix \ref{app:cimpl}. Although directly writing Assembler code
could result in a small performance benefit, this generally increases the work
necessary by an order of magnitude for only limited results. Instruction-level
optimization and in particular register allocation is left to the compiler.
Relying on the compiler mandates a closer study of the generated, optimized
assembler. All source files were compiled using clang version 15.0.7 and
optimization level O3.

@@

Understanding certain choices requires an understanding of the Cortex-A73
instruction pipeline\cite{a72opt:2015}. Being a superscalar processor, it is
able to execute more than one instruction per clock cycle by dispatching
instructions to different execution units working in parallel.

\begin{figure}{h!}
    \centering
    \includegraphics[width=0.7\textwidth]{Figures/a72pipeline.pdf}
    \caption{High-level overview of the Cortex-A72 instruction pipeline}
\end{figure}

@@

Lots of operations require extraction of a certain number of consecutive bits,
usually referred to as a bitfield. Indices for table lookups are generally
attained by right-shifting a larger value stored in a register, then applying
an AND operation to get the lowest $n$ bits and finally writing the result into
the beginning of another register. Due to its usefulness, this operation is
implemented as \texttt{UBFX} for an unsigned bitfield extract and can be used
to implement S-box extraction for subperm lookups which would otherwise take
two or three instructions. Interestingly, the AArch64 instruction set makes
heavy use of instruction aliasing. The logical shift left instruction
\texttt{lsl} for example is an alias of \texttt{UBFX} which itself is an alias
for \texttt{UBFM}.

@@

\subsubsection{Round function}
We will examine the round function in closer detail and compare the source code
with the generated assembly.

@@

\begin{minipage}{.45\textwidth}
    \begin{lstlisting}[language={[ARM]Assembler}, caption={Round function asm}, escapechar=|]
and     v20.16b, v17.16b, v6.16b
add     x9, x20, x8
eor     v20.16b, v20.16b, v16.16b
add     x8, x8, #0x80|\label{lst:cntrincr}|
and     v21.16b, v20.16b, v19.16b
cmp     x8, #0xe70|\label{lst:cntrchk}|
eor     v21.16b, v21.16b, v6.16b
orr     v6.16b, v6.16b, v16.16b
eor     v6.16b, v6.16b, v17.16b
eor     v16.16b, v19.16b, v6.16b
eor     v17.16b, v16.16b, v20.16b
and     v19.16b, v21.16b, v17.16b
eor     v6.16b, v19.16b, v6.16b
and     v19.16b, v7.16b, v4.16b
eor     v19.16b, v19.16b, v5.16b
and     v20.16b, v19.16b, v18.16b
eor     v20.16b, v20.16b, v4.16b
orr     v4.16b, v4.16b, v5.16b
eor     v4.16b, v4.16b, v7.16b
eor     v5.16b, v18.16b, v4.16b
eor     v7.16b, v5.16b, v19.16b
and     v18.16b, v20.16b, v7.16b
eor     v4.16b, v18.16b, v4.16b
    \end{lstlisting}
\end{minipage}\hfill
\begin{minipage}{.45\textwidth}
    \vspace*{7mm}
    \begin{lstlisting}[language={[ARM]Assembler}, firstnumber=24, escapechar=|]
tbl     v18.16b, {v6.16b}, v2.16b
tbl     v19.16b, {v21.16b}, v3.16b
tbl     v21.16b, {v4.16b}, v2.16b
ldp     q6, q4, [x9, #-112]|\label{lst:floatload}|
mvn     v16.16b, v16.16b
tbl     v16.16b, {v16.16b}, v0.16b
tbl     v17.16b, {v17.16b}, v1.16b
mvn     v5.16b, v5.16b
tbl     v5.16b, {v5.16b}, v0.16b
ldp     q22, q23, [x9, #-80]
eor     v6.16b, v6.16b, v16.16b
eor     v16.16b, v4.16b, v17.16b
tbl     v7.16b, {v7.16b}, v1.16b
eor     v17.16b, v22.16b, v18.16b
tbl     v20.16b, {v20.16b}, v3.16b
ldp     q4, q18, [x9, #-48]
eor     v19.16b, v23.16b, v19.16b
eor     v4.16b, v4.16b, v5.16b
ldp     q22, q23, [x9, #-16]
eor     v5.16b, v18.16b, v7.16b
eor     v7.16b, v22.16b, v21.16b
eor     v18.16b, v23.16b, v20.16b
b.ne    197e0
    \end{lstlisting}
\end{minipage}

@@

These kinds of optimizations are pervasive when programming using higher level
languages like C and modern-day compilers more often than not outperform
handcrafted assembly.

@@

An optimized non-SIMD implementation based on the platform-independent
techniques presented in the original paper is relatively easy to achieve since
all functional components are already well defined. One thing to take note of
however is the fact that the specification is based on a big endian
representation while ARMv8 is a little endian machine. The problem of
endianness manifests itself whenever a memory-register interaction takes place,
i.e. for loads and stores. Arithmetic on a register is unaffected since a
register is always treated as one large number with no conception of addresses.

>>>>

Implementations in the C programming language for the presented strategies can
be found in Appendix \ref{app:cimpl}. Although directly writing assembler code
could result in a small performance benefit, this generally increases the work
necessary by an order of magnitude for only limited results. Instruction-level
optimization and in particular register allocation is left to the compiler.
Relying on the compiler mandates a closer study of the generated, optimized
assembler code. All source files were compiled using clang version 15.0.7 at
optimization level -O3.

@@

Understanding certain implementation choices requires an understanding of the
Cortex-A73 instruction pipeline \cite{a72opt:2015}. Being a superscalar
processor, it is able to execute more than one instruction per clock cycle by
dispatching instructions to different execution units working in parallel.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.7\textwidth]{Figures/a72pipeline.pdf}
    \caption{High-level overview of the Cortex-A72 instruction pipeline}
    \label{fig:pipelining}
\end{figure}

@@

Lots of operations require extraction of a certain number of consecutive bits,
usually referred to as a bitfield. Indices for table lookups are generally
attained by right-shifting a larger value stored in a register, then applying
an AND operation to get the lowest $n$ bits and finally writing the result into
the beginning of another register. Due to its usefulness, this operation is
implemented as \texttt{UBFX} for an unsigned bitfield extract and can be used
to implement S-box extraction for subperm lookups which would otherwise take
two or three instructions. Interestingly, the A64 instruction set makes
heavy use of instruction aliasing. The logical shift left instruction
\texttt{lsl} for example is an alias of \texttt{UBFX} which itself is an alias
for \texttt{UBFM}.

@@

\subsubsection{Round function}
We will examine the round function in closer detail and compare the source code
with the generated assembler code.

@@

\begin{minipage}{.45\textwidth}
    \begin{lstlisting}[language={[ARM]Assembler}, caption={\small Round function assembly}, escapechar=|]
and     v20.16b, v17.16b, v6.16b
add     x9, x20, x8
eor     v20.16b, v20.16b, v16.16b
add     x8, x8, #0x80|\label{lst:cntrincr}|
and     v21.16b, v20.16b, v19.16b
cmp     x8, #0xe70|\label{lst:cntrchk}|
eor     v21.16b, v21.16b, v6.16b
orr     v6.16b, v6.16b, v16.16b
eor     v6.16b, v6.16b, v17.16b
eor     v16.16b, v19.16b, v6.16b
eor     v17.16b, v16.16b, v20.16b
and     v19.16b, v21.16b, v17.16b
eor     v6.16b, v19.16b, v6.16b
and     v19.16b, v7.16b, v4.16b
eor     v19.16b, v19.16b, v5.16b
and     v20.16b, v19.16b, v18.16b
eor     v20.16b, v20.16b, v4.16b
orr     v4.16b, v4.16b, v5.16b
eor     v4.16b, v4.16b, v7.16b
eor     v5.16b, v18.16b, v4.16b
eor     v7.16b, v5.16b, v19.16b
and     v18.16b, v20.16b, v7.16b
eor     v4.16b, v18.16b, v4.16b
    \end{lstlisting}
\end{minipage}\hfill
\begin{minipage}{.45\textwidth}
    \vspace*{7mm}
    \begin{lstlisting}[language={[ARM]Assembler}, firstnumber=24, escapechar=|]
tbl     v18.16b, {v6.16b}, v2.16b
tbl     v19.16b, {v21.16b}, v3.16b
tbl     v21.16b, {v4.16b}, v2.16b
ldp     q6, q4, [x9, #-112]|\label{lst:floatload}|
mvn     v16.16b, v16.16b
tbl     v16.16b, {v16.16b}, v0.16b
tbl     v17.16b, {v17.16b}, v1.16b
mvn     v5.16b, v5.16b
tbl     v5.16b, {v5.16b}, v0.16b
ldp     q22, q23, [x9, #-80]
eor     v6.16b, v6.16b, v16.16b
eor     v16.16b, v4.16b, v17.16b
tbl     v7.16b, {v7.16b}, v1.16b
eor     v17.16b, v22.16b, v18.16b
tbl     v20.16b, {v20.16b}, v3.16b
ldp     q4, q18, [x9, #-48]
eor     v19.16b, v23.16b, v19.16b
eor     v4.16b, v4.16b, v5.16b
ldp     q22, q23, [x9, #-16]
eor     v5.16b, v18.16b, v7.16b
eor     v7.16b, v22.16b, v21.16b
eor     v18.16b, v23.16b, v20.16b
b.ne    197e0
    \end{lstlisting}
\end{minipage}

@@

These kinds of optimizations are pervasive when programming using higher level
languages like C and modern-day compilers more often than not outperform
handcrafted assembler code.

@@

An optimized non-SIMD implementation based on the platform-independent
techniques presented in the original paper \cite{camellia:2001} is relatively
easy to achieve since all functional components are already well defined. One
thing to take note of however is the fact that the specification is based on a
big endian representation while ARMv8 is a little endian machine. The problem
of endianness manifests itself whenever a memory-register interaction takes
place, i.e. for loads and stores. Arithmetic on a register is unaffected since
a register is always treated as one large number with no conception of
addresses.

### 2023-05-30 Paper/chapters/introduction.tex
finalize

\chapter{Introduction}
\section{Notation}

@@

\[
    k_7||k_6||\dots||k_1||k_0\leftarrow k_1\ggg 2||k_0\ggg 12||k_7||k_6||\dots||k_3||k_2
\]

@@

With small devices and embedded processors becoming ever more ubiquitous and
essential in areas like consumer electronics or industrial and IoT
applications, the need for low-power, high-performance microprocessors has
increased steadily. With more than 250 billion chips shipped, semiconductors
designed by ARM power 95\% of mobile devices and have found a great many
applications due to their high performance and low power consumption
\cite{armcompany}. The ODROID-N2+ \cite{odroidn2} development board we are
using for testing is powered by an Amlogic S922X processor with four ARM
Cortex-A73 and two weaker ARM Cortex-A53 cores based on the big.LITTLE
architecture combining slower, more power efficient cores with faster, more
power hungry ones. Both these cores are part of the eight generation of ARM
designs known as ARMv8 \cite{armv8:2013} released in 2011 and succeeding the
older 32-bit ARMv7.

ARMv8 defines three architecture profiles for different use cases as well as
dynamic execution states with corresponding instruction sets. This work will
focus on the A profile running in the AArch64 state utilizing the A64
instruction set with NEON and crypto extensions.

>>>>

\chapter{Introduction}

\section{Motivation}

With small devices and embedded processors becoming ever more ubiquitous and
essential in areas like consumer electronics or industrial and IoT
applications, the need for low-power, high-performance microprocessors has
increased steadily. With more than 250 billion chips shipped, semiconductors
designed by ARM power 95\% of mobile devices and have found a great many
applications due to their high performance and low power consumption
\cite{armcompany}. Following the proliferation of these systems, the need for
new and efficient cryptographic algorithms to encrypt and decrypt data arises.

Two promising ciphers are GIFT and Camellia. Both of these have not yet been
implemented for 64-bit ARMv8 - implementations only exist for 32-bit in the case
of GIFT \cite{fixslicing:2020} and Intel AVX-2 and AES-NI in the case of
Camellia \cite{bcfastimplx86:2013}. Our contribution is to fill this gap and
produce efficient implementations of GIFT and Camellia for 64-bit ARMv8.

@@

\[
    k_7||k_6||\dots||k_1||k_0\leftarrow k_1\ggg_2||k_0\ggg_{12}||k_7||k_6||\dots||k_3||k_2
\]

@@

The ODROID-N2+ \cite{odroidn2} development board we are using for testing is
powered by an Amlogic S922X processor with four ARM Cortex-A73 and two weaker
ARM Cortex-A53 cores based on the big.LITTLE architecture combining slower,
more power efficient cores with faster, more power hungry ones. Both these
cores are part of the eight generation of ARM designs known as ARMv8
\cite{armv8:2013} released in 2011 and succeeding the older 32-bit ARMv7.

ARMv8 defines three architecture profiles for different use cases as well as
dynamic execution states with corresponding instruction sets. This work will
focus on the A profile running in the AArch64 state utilizing the 64-bit A64
instruction set with NEON and crypto extensions.
